Locally-Oriented Programming:\\ A Simple Programming Model for Stencil-Based Computations on Multi-Level Distributed Memory Architectures

### 2015-01-24 papers/LOPe/local.tex
Working on paper layout and program models section primarily.

\title{Locally-Oriented Programming: A stunningly simple programming model for GPU programming in Fortran and C++}

\numberofauthors{4}

\author{
% 1st. author
\alignauthor
Craig E Rasmussen\\
       \affaddr{Los Alamos National Laboratory}\\
%       \affaddr{CCS-7, MS B287}\\
       \affaddr{Los Alamos, NM 87545}\\
       \email{crasmussen@lanl.gov}
% 2nd. author
\alignauthor
Matthew J. Sottile\\
       \affaddr{Galois, Inc.}\\
%       \affaddr{421 SW 6th Ave. Suite 300}\\
       \affaddr{Portland, OR 97204}\\
       \email{mjsottile@computer.org}
% 3rd. author
\alignauthor
Dan Nagle\\
       \email{dannagle@verizon.net}
% 4th. author
\alignauthor
Daniel Quinlan\\
       \affaddr{Lawrence Livermore National Laboratory}\\
       \email{dquinlan@llnl.gov}
}

\maketitle

\begin{abstract}
Emerging GPU architectures for high performance computing are well suited to a
data-parallel programming model.  However, programming for these architectures
currently requires learning a new language, e.g., OpenCL or CUDA, or employing
OpenMP compiler directives.  This paper presents preliminary work examining how
programming with a local orientation can be employed in Fortran (or using a
special array library in C++) to provide simple access GPU architectures.  A
locally-oriented programming model is especially useful for stencil-based codes
or algorithms requiring the application of convolution kernels.  In this
programming model, a programmer codes the algorithm {\it only} for a single
array element, but has read-only access to a small sub-array surrounding the
given array element, so that a stencil can be applied.  We demonstrate how a
locally-oriented programming model can be adopted using source-to-source program
transformations in conjuction with a compiler-based infrastructure like ROSE.
\end{abstract}

% A category with the (minimum) three required fields
%\category{D.3.3}{Language Constructs and Features}{Concurrent programming structures}

\input{intro}
\input{programming-model}
\input{shallow-water}
\input{source-to-source}
%\input{static-analysis}
\input{performance}

\section{Conclusions}

The sheer complexity of programming for clusters of many or multi-core
processors with tens of millions threads of execution makes the simplicity of
the data-parallel model attractive.  The increasing complexity of
today's applications (especially in light of the increasing complexity
of the hardware) and the need for portability across architectures
make a higher-level and simpler programming model like data-parallel
attractive.

The goal of this work has been to exploit source-to-source transformations that
allow programmers to develop and maintain programs at a high-level of
abstraction, without coding to a specific hardware architecture.
Furthermore these transformations allow multiple hardware architectures
to be targeted without changing the high-level source.  It also removes the
necessity for application programmers to understand details of the accelerator
architecture or to know OpenCL.

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}
This work was supported in part by the Department of Energy Office of Science,
Advanced Scientific Computing Research.

%%TODO \cite{chamberlain04zpl, roth97stencils}

\bibliographystyle{abbrv}
\bibliography{local}

\balancecolumns
\end{document}

>>>>

\mainmatter  % start of an individual contribution

% first the title is needed
\title{Locally-Oriented Programming:\\ A Simple Programming Model for Stencil-Based Computations
       on Multi-Level Distributed Memory Architectures}

% a short form should be given in case it is too long for the running head
\titlerunning{Locally-Oriented Programming}

\author{Craig Rasmussen%
\thanks{The authors would like to thank Robert Robey and Wayne Weseloh
at Los Alamos National Laborabory for several stimulating
conversations with respect to programming models.}%
\and Matthew Sottile\and Dan Nagle\and Soren Rasmussen}

\authorrunning{Locally-Oriented Programming}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{441 McKenzie Hall,\\
5246 University of Oregon, Eugene, Oregon, USA\\
\mailsa\\
\mailsb\\
\url{https://openfortran.uoregon.edu}}

%
% NB: a more complex sample for affiliations and the mapping to the
% corresponding authors can be found in the file "llncs.dem"
% (search for the string "\mainmatter" where a contribution starts).
% "llncs.dem" accompanies the document class "llncs.cls".
%

%\toctitle{Lecture Notes in Computer Science}
%\tocauthor{Authors' Instructions}
\maketitle

\begin{abstract}

Emerging hybrid accelerator architectures for high performance computing are often well suited for
usage of a data-parallel programming model.  However, programmers for these emerging architectures
face a steep learning curve that frequently requires learning a new language, e.g., OpenCL or CUDA,
or employing OpenMP compiler directives.  Furthermore, the distributed (and frequently multi-level)
nature of the memory organization of clusters of these architectures provides an additional level of
complexity for the programmer.  This paper presents preliminary work examining how programming with
a local orientation can be employed to provide simple access to accelerator architectures.  A
locally-oriented programming model is especially useful for stencil-based codes or algorithms
requiring the application of convolution kernels.  In this programming model, a programmer codes the
algorithm by modifying \emph{only a single array element}, but has read-only access to a small
sub-array surrounding the given array element, so that a stencil can be applied.  We demonstrate how
a locally-oriented programming model can be adopted as an embedded, domain-specific language using
source-to-source program transformations.

\keywords{domain specific language, stencil compilers, distributed memory parallelism}
\end{abstract}

\input{motivation}
\input{programming-model}
\input{appendix-a}

\subsection{Acknowledgments}
This work was supported in part by the Department of Energy Office of Science,
Advanced Scientific Computing Research.

\end{document}

### 2015-01-24 papers/LOPe/motivation.tex
Working on paper layout and program models section primarily.

\subsection{Coarray Fortran Comparison}

Below is the motivational example from the original Numrich and Reic paper.
{NOTE: This should sortof/kindof show how it would work between nodes?  Next
section shows OpenMP example for parallelism at the thread level.}

{\small
\begin{verbatim}

subroutine laplace (nrow,ncol,u)
   integer, intent(in) real, intent(inout) real integer
   ! that refers to the current image
   left = me-1
   if (me == 1) left = ncol
   right = me + 1
   if (me == ncol) right = 1
   call sync_all( (/left,right/) ) ! Wait if left and right
   ! have not already reached here
   new_u(1:nrow)=new_u(1:nrow)+u(1:nrow)[left]+u(1:nrow)[right]
   call sync_all( (/left,right/) )
   u(1:nrow) = new_u(1:nrow) - 4.0*u(1:nrow) end subroutine laplace

\end{verbatim}
}

Below is the equivalent implementation in LOPe.

{\small
\begin{verbatim}

pure CONCURRENT subroutine laplace(U)
   real, intent(inout), HALO(-1:*:1,-1:*:1) :: U(0:,0:)

   U(0:0) = U(-1,0) + U(+1,0) + U(0,-1) + U(0,+1) - 3*U(0,0)

end subroutine laplace

\end{verbatim}
}

NOTE: Following not a criticism of CAF as CAF is a general purpose parallel
programming language and LOPe only pertains to the halo parallel pattern.

\subsubsection{Improvements}
\begin{itemize}

\item
LOPe requires that the implementation of the algorithm to be separate from the
call to effect the halo transfer.  Removing boundary condition specification
(e.g., the cyclic boundary conditions implemented in the CAF example) from the
algorithm allows the boundary conditions to be changed and without changing
algorithm code.

\item
LOPe applies the transfer of halo memory across multiple (possibly) levels of
memory with the intrinsic {\tt transfer\_halo} function (not shown here).  Thus
the LOPe algorithm can be run on a machine with many interconnected nodes, each
containing hybrid processor cores.  The coarray example can only be run on
multiple nodes (called images in Fortran) without accelerator cores.

\item
The algorithm implementation is separate from user-specified synchronization,
e.g., {\tt call sync\_all}.  In LOPe, synchronization is subsummed in the
semantics of the CONCURRENT attribute and the {\tt transfer\_halo} function.

\item
The algorithm implementation is separate from any specification as to where the
array memory is located.  The CAF example explicity denotes where memory is
located with the {\tt [left]} and {\tt [right]} syntax where left and right
specifiy a processor topology.

\item
The algorithm implementation is separate from any specification as to where the
algorithm is to be executed.  The CAF example explicity denotes where a statement is to
executed with control flow construct like {\tt if (me == 1)}.

\item
The LOPe implementation is easier to understand and frequently follows the
mathematical algorithm directly.  For example, the CAF example adds 4 neighbors
plus the center value to make the implementation with direct remote coarray
access possible, while the LOPe example is able to implement the same algorithm
with fewer operations by adding 4 neighbors (not including the center array
element) and then only subtracting 3 center values.

\item
The semantics of LOPe makes explicit management of array temporaries (e.g., {\tt
  u} and {\tt new\_u} by programmers unnecessary (though still possible).
Because in LOPe the halo region is a language construct, the compiler is better
able to manage temporary buffers than users on the target hardware platform.

\end{itemize}

\subsubsection{Errors that are constrained by the language}
\begin{itemize}

\item
A programmer is not able to store data to the halo region.  If this were
allowed, one thread could overwrite another threads data at undefined times.
The compiler is able to catch this class of error.

\item
A programmer can't make indexing errors in a concurrent routine by going out of
bounds of the array plus halo memory.  The compiler is able to catch this class of
error at compile time unless the halo shape is assumed (e.g., HALO(:,:)).

\item
A programmer is not able to cause race conditions by forgetting to create and
use temporary arrays properly.  In LOPe it is the comilers responsibility to
store data in temporary memory.

\item
A programmer can't make synchronization errors as synchronization is implicit in
the CONCURRENT attribute.  A thread running a CONCURRENT procedure is provided
with a copy of it's array element (plus halo) that is consistent with the state
of memory at the time of invocation of procedure.  Stores to an individual
thread's array element (by that thread) is never visible to other threads.
It is possible to relax this restraint with a halo synchronization intrisic.
However, LOPE encourages the creation of small functions and lets the compiler
fuse the procedures together and provide the necessary synchronization.

\end{itemize}

Please note that this example is somewhat unfair because in practice CAF is usually
refactored in a locally oriented way by so that communication and
synchronization separated into separate procedures.  Locally-oriented programming should be viewed as programming methodology with LOPe as a particular instance.

This is the normal way that large complex MPI and CAF programs are implemented, i.e., using
halo cells.  LOPe proposes to formalize this common pattern into the Fortran
language providing the compiler information in order to spread computation over
more hardware resources and reduce complexity for the programmer.

\subsection{OpenMP Comparision}

The coarray Fortran example shows how LOPe

>>>>

The concepts in this paper were motivated by problems encountered during
the development of PetaVision, a C++ software framework designed for simulating
large networks of spiking neurons in the visual cortex of primates.  PetaVision
was designed to run on massively parallel hardware architectures and a primitive
version of PetaVision achieved over a Petaflop/s of sustained single precision
performance on Roadrunner, the fastest computer in the world for a brief
period of time.

TODO: Describe neural layers, spiking neurons, synaptic connections, weights,
and plasticity. Cite Heubel.

TODO: Give a simple 1D (cartoon) example of connections within a neural network in
Figure 1.  Describe the size of visual cortex in terms of the number of neurons,
the number of synaptic connections, and performance.  Reference SC Gorden Bell
paper from IBM.

### 2015-01-24 papers/LOPe/programming-model.tex
Working on paper layout and program models section primarily.

@@

\begin{verbatim}
  pure CONCURRENT subroutine average(A)
     real, HALO(:,:) :: A

     A(0,0) =              A(+1,0)                &
               + A(0,-1) + A( 0,0)   +  A(0,+1)   &
                         + A(-1,0)

     A(0,0) = (1./5.) * A(0,0)

  end subroutine
\end{verbatim}

@@

This section describes the hierarchical memory layout and how memory consistency between the 3-levels of memory is maintained.  It describes what is the compiler's responsibility and which is the user's responsibility.

>>>>

The LOPe (for Locally-Oriented Programming extensions) programming model restricts the programmer ---
when implementing a stencil algorithm --- to a local view of the array index space.  Within a LOPe
function, only the local array element (plus a small halo region surrounding the local element) is
visible to the programmer.  This serves to reduce complexity by removing all boundary conditions and
processor topology information from the implementation of the algorithm.

%%This reduction in complexity reduces programming errors.  While developing the convolution example described later in the paper we made an indexing error in applying the 2D stencil loops in the standard serial Fortran test implementation.  This error required over 3 hours of programming time to repair.  First the error had to be isolated to the function implementing the convolution (it was first thought to be in the complicated tiff image output routine as the convolution code was ``thought'' to be too simple to wrong.  Then the index error has to be understood.  As will be seen, LOPe makes it more difficult to make these errors as Fortran array intrics can be used.  In some instance, the restricted semantics of LOPe allows the compiler to catch errors (e.g., some errors involving race conditions).

LOPe was implemented as an embedded domain specific language (DSL).  Fortran was chosen as the base
language for LOPe, although in principle, the same techniques could be applied to language such as C
or C++ (in conjuction with some form of array extensions such as those proposed by Intel).  Fortran
was chosen because it already provides a rich array-based syntax.  Readers who are unfamiliar with
Fortran syntax may wish to consult Appendix A for a brief description of Fortran notation.

\subsection{LOPe Syntax Extensions}

There are only a few syntax additions required for LOPe.  In the code examples describing these
additions (and throughout this paper as well), DSL syntax additions are highlighted by the usage of
capitalization for all new keywords.

\subsubsection{HALO.}

The principle semantic element of LOPe is the concept of a halo region.  A halo is an ``artificial''
or ``virtual'' region surrounding an array that contains boundary-value information.  Halo (also
called ghost-cell) regions are commonly employed to unify array indexing schemes in the vicinity of
an array boundary, so that the array may be referenced using indicies that fall ``outside'' of the
domain of the array.  In LOPe, the halo region is given explicit syntax so that the compiler can
exploit this knowledge.  For example, a halo region can be explicitly provided in a declaration
statement of the form,
\begin{verbatim}
  real, allocatable, dimension(:), HALO(1:*:1) :: A
\end{verbatim}
This statement indicates that array \texttt{A} is one dimensional, will be allocated later when it
is provided with a specific size, and has a halo region of 1 surrounding the array on both sides.
The notation \texttt{M:*:N} is to be read as describing a halo region of size \texttt{M} to the
left, size \texttt{N} to the right, and with an arbitrary size in the middle (denoted by the $*$
symbol).  When used to describe a formal parameter of a function, the \texttt{HALO} keyword may
``assumed'' (from the actual argument by the compiler) as is the following statement,
\begin{verbatim}
  real, HALO(:,:) :: U
\end{verbatim}
which in this instance indicates that \texttt{U} is a two-dimensional array, with a halo region
whose size is ``assumed'' by the compiler from the explicit halo size of the actual array argument
(provided at the calling site of the function).  In LOPe, there is no need in this instance for a
repetitive \texttt{dimension(:,:)} specification as it is inferred by the \texttt{HALO}
specification.

\subsubsection{CONCURRENT.}

The second keyword employed by LOPe (although in this case extended) is \texttt{CONCURRENT}.  The
\texttt{concurrent} keyword already exists in the form of a \texttt{do concurrent} loop construct (a
Fortran 2008 feature).  By using \texttt{do concurrent}, the programmer asserts that specific
iterations of the loop body may be executed by the compiler in \emph{any order;} even
\emph{concurrently.}  For example, the following,
\begin{verbatim}
  pure CONCURRENT subroutine Laplace(U)
     real, HALO(:,:) :: U
     U(0,0) =                 U(0,+1)               &
              +  U(-1,0)  - 3*U(0, 0)  +  U(+1,0)   &
                          +   U(0,-1)
  end subroutine Laplace
\end{verbatim}
defines a function that can be used as part of an iterative solution of Laplace's equation in two
dimensions.  A function with the \texttt{pure} and \texttt{CONCURRENT} attributes may be called from
with a \texttt{do concurrent} loop.  One should imagine that this function is called \emph{for each}
\texttt{(i,j)} iterate of the loop.

\subsubsection{LOPe index notation.}
In the \texttt{Laplace} example shown above, the \texttt{(0,0)} array element is the \emph{local} 
array element and only the local element may be modified.

\input{coarray-comparison}

@@

@@

This section describes the hierarchical memory layout and how memory consistency between the 3-levels of memory is maintained.  It describes what is the compiler's responsibility and what is the programmer's responsibility.
